---
title: "Classifying legendary pokemon"
---

```{r setup}
library(pokedex)
library(tidyverse)
library(GGally)
```

## Key stats

Pokemon have some key stats, lets look at how they correlate and are distributed

```{r}
ggpairs(
  pokemon %>%
    select(
      base_experience,
      hp,
      attack,
      defense,
      special_attack,
      special_defense,
      speed
    ),
  lower = list(continuous = "density"),
  progress = FALSE
)
```

Base experience corellates the most strongly with all the 6 other stats
has 3 distinct groups

Special defense and defense correlate well

Negative corellation between speed and defense
  + Weak corelation between speed and hp and speed and special defense

## Is this evolution?

Lets make a (rough) estimate of how far down an evolutionary line a creature is


```{r}
pokemon %>% 
  filter(identifier == "eevee" | evolution_chain_id == 67)
```

```{r}
pokemon %>% filter(is.na(evolution_chain_id))
  group_by(evolution_chain_id) %>%
  summarise( n = n())
```


```{r}
ggpairs(
  pokemon_plus %>%
    select(
      evo_steps,
      base_experience,
      hp,
      attack,
      defense,
      special_attack,
      special_defense,
      speed,
      stats_total
    ),
  lower = list(continuous = "density"),
  progress = FALSE
)
```

* generally stats increase up the evo steps
* base-exp has gap, then the top figure, legendary pkmn
* hp also closely mirrors, higher level = more damage to deal with

## Split point for  legendary pokemon

```{r}
ggplot(pokemon_plus, aes(base_experience, evo_steps)) +
  geom_point()
```

```{r}
pokemon_plus %>% 
  filter(base_experience > 200 & evo_steps == 1) %>% 
  arrange(stats_total)
```

