---
title: "Tidy Tuesday and space to learn"
output: github_document
---

[TidyTusdays](https://github.com/rfordatascience/tidytuesday) are a weekly R Community event where people learn about RStats by practising with a different data set each week. Last week the [Cardiff R User group](https://www.meetup.com/Cardiff-R-User-Group/) worked on the volcanoes data-set :volcano:, and something in it really tripped me up. We've done this a few times now, and are building up our work in [this GitHub repo](https://github.com/CaRdiffR/tidy_thursdays).

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
volcano <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-05-12/volcano.csv')

```

## Data structure

Let's have a look at what the data is

```{r}
volcano %>% 
  str()
```


## Objective 

Lots of character columns, and some with some slightly funky formatting, such as `/` and variations on a theme with `(s)` and `(es)`. We've also got a bunch of 'sparse' data in the columns that start with `major_rock` or `minor_rock`. That look like spaces. R has a rich set of tools for dealing with effectively missing data a little more effectively, so lets clean this up by setting the missing data to be explicit `NA`. In this case, as the column is a character type, we need to `NA_character` to fill it up.


## Failing solution

```{r}
volcano %>%
    mutate_at(
    .vars = vars(starts_with(c("major_rock", "minor_rock"))),
    .funs = ~ case_when(
      . == " " ~ NA_character_,
      TRUE ~ .
    )
  ) %>% 
  select(starts_with(c("major_rock", "minor_rock"))) %>% 
  head() %>% 
  knitr::kable()
```

Well, that doesn't quite work. What I want is to have the blank spots filled up with `NA`. Is it my code? It's not the most basic solution, using some of the tidyeval concepts such as `vars` and `funs`. Lets make it as simple as possible.

```{r}
volcano %>% 
  filter(major_rock_5 == " ") %>% 
  knitr::kable()
```

Odd. I can't find any values that are just spaces, even though they are printed out that way! I know they are there, I can see them! Luckily, the point of these projects is to learn, and to learn from each other in the group :school:.

## Problem

My buddy [Heather](https://twitter.com/HeathrTurnr) was I think a little surprised when I demonstrated this, but within a few minutes she'd worked it out. It's encoded as a _non_breaking space_. She linked this blog in our chat about [non-braking spaces](https://blog.tonytsai.name/blog/2017-12-04-detecting-non-breaking-space-in-r/) and offered us the cryptic solution:

```
"\u00A0"
```

The bog goes into detail about what this is but the tl;dr is: "It looks like a space but it's not and it's [designed that way](https://en.wikipedia.org/wiki/Non-breaking_space)".

So a quick tweak to the code and...

## Functional solution

```{r}
volcano %>%
    mutate_at(
    .vars = vars(starts_with(c("major_rock", "minor_rock"))),
    .funs = ~ case_when(
      . == "\u00A0" ~ NA_character_,
      TRUE ~ .
    )
  ) %>% 
  select(starts_with(c("major_rock", "minor_rock"))) %>% 
  head() %>% 
  knitr::kable()
```


## Full solution

Success! After digging around a little, I discovered `str_trim` and `str_squish` can be used for this as well to make a perfectly tidy solution!

```{r}
volcano %>%
    mutate_at(
    .vars = vars(starts_with(c("major_rock", "minor_rock"))),
    .funs = ~ case_when(
      str_trim(.) == "" ~ NA_character_,
      TRUE ~ .
    )
  ) %>% 
  select(starts_with(c("major_rock", "minor_rock"))) %>% 
  head() %>% 
  knitr::kable()
```

Let's try and breakdown what is happening in this solution by concept, and then outline the routine in human language to finish.

## Concepts

### Non-breaking spaces

* the 'missing' values are not real spaces, they are _non-breaking spaces_.
* [`stringr::str_trim`](https://stringr.tidyverse.org/reference/str_trim.html) and `stringr::str_squish` removes space from either the ends or all the way through a character string depending on what else is happening in the string and what you need from the solution.
* [`mutate_at`](https://dplyr.tidyverse.org/reference/mutate_all.html) is a buddy of `mutate`, where you use _functional programming_ style to apply a function over a collection of columns.
* this means that for the context of evaluation, we will be getting `""`, where as previously we were seeing `" "` which was actually encoded as `"\u00A0"`

### Variable selection

* [`starts_with`](https://dplyr.tidyverse.org/reference/select.html#useful-functions) is a select helper that is designed for cases when a related value is stretch wide across multiple columns with similar names, and returns a vector of column names filtered to your criteria.
* [`vars`](https://dplyr.tidyverse.org/reference/vars.html) automatically _quotes_ the names of the columns to _evaluate_ later in context and is almost always used as a wrapper to the `.var =` argument when it's supported by a function.
* this means that we will be doing on operation on each of the columns selected.

### Formula function

* `.funs =` argument, like `.var =`, has a counterpart [`funs()`](https://dplyr.tidyverse.org/reference/funs.html), but this is being deprecated in favour of the _expression notation_.
* [`case_when`](https://dplyr.tidyverse.org/reference/case_when.html) is an alternative version of the more common `if else` operation. 
* `~` is a special operator that is key to the expression notation. It effectively separates the Left Hand Side (LHS) of an expression from the Right Hand Side (RHS). It's used in two ways in this code. `case_when` uses full expressions to represent what should happen on the RHS when the criteria of the LHS is met. `.funs =` uses it to make a lambda style formula. This is sometimes referred to as a [_quosure_](https://dplyr.tidyverse.org/articles/programming.html#quoting). 
* `.` is also a special operator, and I recommend reading the [documentation of pipe](https://magrittr.tidyverse.org/reference/pipe.html) `%>%`. The idea of it here is to reference the data being operated on itself. In this specific case, it's each value from each of the selected columns for equivalence to an empty space.

## Step-by-step 

Did you follow all that? It's a minefield I know, but it allows us to do something very powerful in only a few lines. As an alternative way of understanding what this does, here's the step by step:

1) Get all the columns that start with either `"major_rock'` or `"minor_rock"`
1) For each of the those columns trim any value at the start or end that is whitespace, including non-breaking white space temporarily, without modifying the underlying data
1) If the value after that is an empty string, replace it in the same column with the `NA` value for characters
1) If the value after does not pass that test, use the original value

So do you _have to_ program this way? No, not really. You could manually create the list of columns you want to modify, but that would be prone to human error and what if you end up with `"minor_rock_6"` or `"major_rock_100"`? You could always make a traditional `if else` structure, but that can get long fast if there are multiple conditions to check for. How about using the character string `"\u00A0"` to test for equivalence? Well, that would work now, but how do you pick up if they change suddenly to actual spaces? Or another whitespace encoding? Programming like this keeps code readable, maintainable, and robust. 

Is this overkill for tidy Tuesdays? Yes, absolutely. Are the problems that you solve with this approach purely academic? No, not at all. Plus, doing all that work in 129 characters is pretty neat. Excluding spaces.