---
title: "Predicting pokemon power"
description: "a description"
series: "a series"
tags: ["tag1", "tag2"]
output: github_document
---

```{r setup, message = FALSE}
library(pokedex)
library(tidyverse)
```

```{r}
pokemon %>% 
  group_by(generation_id) %>%
  skimr::skim_without_charts()
```
## How strong is any given Pokemon?

Pokemon games have a very familiar cycle. You start with one of 3 Pokemon. You adventure out with your new buddy, facing tougher and tougher Pokemon, in greater variety. Many of your Pokemon evolve over time, and eventually you find the end-game legendaries in a climactic battle of titans! 

You can even see this story in the data. Here is the `base_experience` of all the Pokemon.

```{r}
pokemon %>% 
  ggplot(aes(x = id, y = base_experience, fill = generation_id)) +
geom_point()
```



```{r}
pokemon %>%
  mutate(id = as_factor(id)) %>% 
  select(
    name,
    id,
    generation_id,
    base_experience,
    hp,
    attack,
    defense,
    special_attack,
    special_defense,
    speed
  ) %>%
  pivot_longer(where(is.numeric), names_to = "stat", values_to = "value") %>%
  ggplot(aes(x = id,
             y = value,)) +
  geom_point() +
  facet_grid(stat ~ generation_id, scales = "free")
```


