---
title: "Introducing the pokedex R package"
output: github_document
---

I made an R data package to make Pokemon data more usable in R! 

```{r setup}
library(pokedex)
library(tidyverse)
library(knitr)
```

# How many Pokemon in the package?

I've tried to make the data set 'tidy' from the start, so we can use `summarise` to count them, and `kable` to make some dev.to friendly markdown tables.

```{r total_count}
pokemon %>% 
  summarise(count = n()) %>% 
  kable()
```

# Types

Types are pretty key to Pokemon. Lets have a quick look at the Kanto starters and types.

```{r kanto_table}
pokemon %>% 
  top_n(n = -9, wt = species_id) %>% 
  select(identifier, type_1, type_2) %>%
  kable()
```

# Single and dual types

So Pokemon can have either 1 or 2 types. What's the split between single type and dual type Pokemon?

```{r count_of_dual_and_single_types}
pokemon %>%
  mutate(dual_type = case_when(is.na(type_2) ~ TRUE,
                               TRUE ~ FALSE)) %>%
  group_by(dual_type) %>%
  summarise(count = n()) %>% 
  kable()
```

So, it's nearly a 50:50 split of Pokemon that are single type to Pokemon that have 2 types.

# How many by type?

But there are also quite a few types of Pokemon. Starting with the primary type, lets make a quick chart to understand the distribution of primary types. Using `group_by` will mean the `summarise` gets calculated _per group_. We can then pipe directly into `ggplot` for a col chart with `geom_col`.

```{r primary_type}
pokemon %>%
  group_by(type_1) %>%
  summarise(count = n()) %>%
  ggplot(aes(x = type_1, y = count)) +
  geom_col() +
  labs(title = "Pokemon by primary type")
```
 
Lots of water type Pokemon, and lot's of normal type Pokemon, but very few flying types. Interesting. How about the secondary types?

```{r secondary_type}
pokemon %>% 
  filter(!is.na(type_2)) %>% 
  group_by(type_2) %>% 
  summarise(count = n()) %>% 
  ggplot(aes(x = type_2, y = count)) +
  geom_col() +
  labs(title = "Pokemon by secondary type",
       caption = "For Pokemon with dual type")
```

Look at all those 'mons with flying as a secondary type! The thing is that, game-wise, the _order_ of the typing doesn't matter. We can easily count the occurrence of a specific type in either primary or secondary position with `pivot_longer`.

`pivot_longer` is actually a newer tidyverse function. It is complemented with `pivot_wider` and this pair are intended to eventually replace `spread` and `gather`. By filtering out the `NA` I remove any observations of secondary types for Pokemon that don't actually have them.

```{r either_type}
pokemon %>% 
  select(identifier, type_1, type_2) %>% 
  pivot_longer(-identifier, names_to = "slot", values_to = "type") %>% 
  group_by(type) %>% 
  summarise(count = n()) %>% 
  filter(!is.na(type)) %>% 
  ggplot(aes(x = type, y = count)) +
  geom_col() +
  labs(title = "Pokemon by either type",
       caption = "This will count a dual type Pokemon twice,\nonce for each type")
```

So is there any consistency in order at all?

```{r ghost_fire_is_a_fire_ghost}
pokemon %>%
  filter((type_1 == "ghost" & type_2 == "fire") |
           (type_1 == "fire" & type_2 == "ghost")) %>% 
  select(identifier, type_1, type_2)
```

It doesn't look like it. a `ghost fire` Pokemon and a `fire ghost` Pokemon both turn up. I'd like to see what the coincidence rate is of each type in dual type Pokemon, so I need to get some ordering in. I can use `case_when` in `mutate` to create a two new columns in the data. I can make 2 in one call because `mutate` supports multiple _expressions_, each of which names a column, and then operates conditionally on the other 2 type columns. These new columns will:

* always have a value in the column `type_1_ordered`
* if the Pokemon is dual type, have a value in `type_2_ordered`
* always have the types alphabetically ordered between the two columns. i.e. it will always be `fire, ghost`, never `ghost, fire`.

```{r consistent_typing}
pokemon %>%
  mutate(
    type_1_ordered = case_when(is.na(type_2) ~ type_1,
                               type_1 < type_2 ~ type_1,
                               TRUE ~ type_2),
    type_2_ordered = case_when(type_1 > type_2 ~ type_1,
                               TRUE ~ type_2)
  ) -> pokemon
```

What might the distribution be of the flying secondary type, per primary type?

```{r count_dual_type}
pokemon %>%
  mutate(type_combined = case_when(
    !is.na(type_2_ordered) ~ paste(type_1_ordered, type_2_ordered),
    is.na(type_2_ordered) ~ type_1_ordered
  )) %>%
  group_by(type_combined) %>%
  summarise(count = n()) %>%
  arrange(desc(count)) %>%
  mutate(type_combined = as_factor(type_combined)) %>%
  ggplot(aes(x = type_combined, y = count)) +
  geom_col() +
  labs(title = "Count of Pokemon by dual type",
       caption = "Ordered by count") + 
  theme(axis.text.x = element_text(angle = 90))
```

So the most often occurring dual type is flying normal. That explains the first 2 charts. It's a bit tricky to see the rest though. Lets make a more useful plot.

```{r covariance_by_type}
pokemon %>% 
  group_by(type_1_ordered, type_2_ordered) %>% 
  filter(!is.na(type_2_ordered)) %>% 
  summarise(count = n()) %>% 
  ggplot(aes(x = type_1_ordered, y = type_2_ordered, size = count)) +
  geom_point() +
  labs(title = "Coincidence of a particular dual type")
```

So `flying normal` has the biggest count, with there being quite a few `bug flying`. That makes sense, as so many bug Pokemon have wings! There are also lot's of `bug poison` and `grass poison`. That makes sense too, as so many bugs and plants are poisonous! How many Pokemon have unique types though?

```{r count_unique_dual_types}
pokemon %>% 
  group_by(type_1_ordered, type_2_ordered) %>% 
  filter(!is.na(type_2_ordered)) %>% 
  summarise(count = n()) %>% 
  filter(count == 1) %>% 
  ungroup() %>% 
  summarise(count = n()) %>% 
  kable()
```

24 Pokemon have unique dual types. Out of 807 that isn't very many! Maybe these Pokemon might be particularly useful? I'll try and work it out...

# Game Over

This post has been a simple example of both the data in the package, but also the `tidyverse` methods of doing Exploratory Data Analysis. You can find out more about tidyverse [here](https://www.tidyverse.org/)

I got the raw data from [this repo by veekun](https://github.com/veekun/pokedex). My package is available [here](https://github.com/DaveParr/pokedex), and the particular version I used for this post is [here](https://github.com/DaveParr/pokedex/commit/67638e8bc52d58bb0c38534b7c2acc9a78b42053). Though it's in a pretty raw state, I hope to improve over time. 

I made this package to have a bigish, diverse set of data to play with, that lots of people recognise, and that has some inherent real world application. Pokemon is a huge franchise with multiple instalments. Lots of people have played it, and even if you haven't you probably have an intuition about what a Pokemon is, and what data about a Pokemon might make sense, and mean in context with other Pokemon. Feel free to fork and mess around with as you like. I hope its fun, and maybe even useful!

